---
title: 'Lab 3: Exploring Happiness'
subtitle: "Due: Thursday 10/03/2019 by 10 pm"
author: "Yealim Sung and Elcy Timothy"

date: "09/26/2019"
output: html_document
---


## Background

The General Social Survey (GSS) has been run by NORC every other year since 1972 to keep track of current opinions across the United States. We are working with an excerpt of the dataset that covers years 1972 to 2016. The variable names along with a brief description are as follows: 

| Variable      | Description   |    
| ------------- |:-------------:| 
| Year          | Year for this respondent | 
| Age           | Age of respondent | 
| Sex           | Sex of respondent     | 
|....           | .....      | 

Now it's important to understand what each column stands for. You can find a codebook with explanations for each of the variables at https://gssdataexplorer.norc.org/

## Data Cleaning 

Initial exploration revealed that some data processing was needed before we could proceed with further visualization. All missing values were encoded with 'NA' as follows:
```{r}
#insert relevant R code for Q1 here
happy <- readRDS("HAPPY.rds")
names(happy) = tolower(names(happy))
happy[happy == "IAP"] <- NA
happy[happy == "DK"] <- NA
clean_happy <- na.omit(happy)
```

In total there were `25170` number of missing values for the following `62376` variables. 

Variables were also cast into the correct type in order to ease data visualization. This was done by:
```{r}
#insert relevant R code for Q2 here
clean_happy[clean_happy == "89 OR OLDER"] <- as.numeric(89)
head(clean_happy)
```
Setting the string "89 OR OLDER" as numeric value 89 and assuming it will be 89 so that all the values in column age will be numeric.

Factors were also organized into meaningful order as follows: 
```{r}
#insert relevant R code for Q3 here
sortedByMarital <- clean_happy[order(clean_happy$marital),]
sortedByAge <- sortedByMarital[order(sortedByMarital$age),]
head(sortedByAge)
```
First we sorted the dataset by marital status and then sorted that dataset by age.


## Data Exploration 

Two variables of interest are `sex` and `partyid`. Their distributions are as follow: 
```{r}

library(ggplot2)
ggplot(sortedByAge, aes(sex, fill=partyid)) + geom_bar() + labs(x = "Sex", y = "Political Party ID", title = "Distribution of Political Party between the Genders")

ggplot(sortedByAge, aes(sex, partyid)) + geom_jitter() + labs(x = "Sex", y = "Political Party ID", title = "Jitter plot on distribution of Political Party")
 
```
We chose variables `sex` and `partyid` because one of us is social science major and we wanted to see how one's political view affects the happiness and which sex are more interested in what political party. It seems like female are more interested in politics from this dataset. The most popular party id was 'NOT STR DEMOCRAT'. We will explore more on how it affects happiness on the next section.

```{r}
ggplot(sortedByAge, aes(happy, fill=partyid)) + geom_bar() + labs(x = "Happiness", y = "Political Party ID", title = "Bar chart of happiness over political party id")

ggplot(sortedByAge, aes(happy, partyid)) +geom_point(aes(color=sex)) + geom_smooth() + labs(x = "Happiness", y = "Political Party ID", title = "Happiness over political party id")
```

Comparing the variable party ID to happiness. We found that those who identity as NOT STR DEMOCRAT seem to have the largest distribution of happiness at pretty happy. The second party ID with the biggest distribution of happiness is STRONG DEMOCRAT at pretty happy.In this data set the repspondents are at the level of PRETTY HAPPY more than the others. And the party ID that is happier overall is NOT STRE DEMOCRAT and  STRONG DEMOCRAT in both categories of PRETTY HAPPY and VERY HAPPY. 






